Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-23 sigcse-17/intro.tex
sigcse: import text on 410 from edupar, with some violence
sigcse: add some pretty pictures

% TODO: lol.
Building real-world multithreaded concurrent programming systems is well-known to be notoriously difficult to design, implement, test, debug, hard to get right, challenging, and error-prone arguably because of the difficulties of finding, reproducing, and fixing concurrency errors, attributable to the astronomically large number of thread interleavings, which arise at random during normal program execution.

@@

Incorporating SMP into the \fourten curriculum would expand the challenges students face in a single semester.
However, we wish to avoid compromising the complexity
%, hardware fidelity,
and learning experience of the project set.
One solution would be to ease the burden of debugging concurrency errors.
Hence, we have built \landslide%
% TODO camready
%\cite{landslide}
, a tool that uses stateless model checking \cite{verisoft,dpor,chess-icb}
as an alternative to the conventional stress tests.
%to force kernel threads to take execution paths that expose concurrency bugs.
\landslide can test student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
We present our experience sharing \landslide with volunteering students who used it to find bugs in their own code,
and discuss the possibility for integrating it as a main component of operating systems curricula.

>>>>

%Building real-world multithreaded concurrent programming systems is well-known to be notoriously difficult to design, implement, test, debug, hard to get right, challenging, and error-prone arguably because of the difficulties of finding, reproducing, and fixing concurrency errors, attributable to the astronomically large number of thread interleavings, which arise at random during normal program execution.

@@

Incorporating SMP into the \fourten curriculum would expand the challenges students face in a single semester.
However, we wish to avoid compromising the complexity
%, hardware fidelity,
and learning experience of the project set.
One solution would be to ease the burden of debugging concurrency errors.
Hence, we have built \landslide%
% TODO camready
%\cite{landslide}
, a tool that uses stateless model checking \cite{verisoft,dpor,chess-icb}
as an alternative to the conventional stress tests.
%to force kernel threads to take execution paths that expose concurrency bugs.
\landslide tests student thread libraries fully automatically, and can test kernels provided some manual instrumentation.
We present our experience sharing \landslide with volunteering students who used it to find bugs in their own code,
and discuss the possibility for integrating it as a main component of operating systems curricula.

### 2016-08-23 sigcse-17/landslide.tex
sigcse: add some pretty pictures

Motivated by the stress tests' inconsistency at finding concurrency bugs,
% TODO camready
we have built Landslide%\cite{landslide}
, a stateless model checker for \pebbles kernels and thread libraries.
Stateless model checking \cite{verisoft} offers an alternative
%concurrency testing approach
to stress testing,
in which the testing framework controls scheduling to force a new thread interleaving to execute in each iteration of a test.
It follows in the footsteps of related tools such as CHESS \cite{chess},
but also provides several important features for use in an educational setting,
such as heuristics to handle unusual student solutions, kernel-level testing, and user-friendly progress reports and HTML debugging traces.

@@

{\bf Example.}
Figure \ref{fig:paradise}(a) shows an example concurrency bug which we will use as a running example.
The code implements a producer/consumer program, in which one thread sends data to the other through a shared list,
synchronized with a mutex and condition variable.
However, because the consumer code only checks the list size once before blocking,
multiple simultaneous consumers can race,
causing a segfault when one receives a null \x{data} pointer. %and tries to dereference it

@@

The branching possibilities arising from each preemption point produce an execution tree, or state space,
as shown in Figure \ref{fig:paradise}(b).
%As these state spaces are exponentially-sized,
Landslide uses Dynamic Partial Order Reduction \cite{dpor} to identify and prune equivalent interleavings
to mitigate the exponential explosion of these state spaces.
% TODO make sure this gets mentioned later, if not here
%and state space estimation \cite{estimation} to predict which
% TODO CAMREADY: cite oopsla
% In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.

@@

% TODO: make fig trace

@@

>>>>

Motivated by the stress tests' inconsistency at finding concurrency bugs,
% TODO camready
we have built Landslide%\cite{landslide}
, a stateless model checker for \pebbles kernels and thread libraries.
Stateless model checking \cite{verisoft} offers an alternative
%concurrency testing approach
to stress testing,
in which the testing framework controls scheduling to force execution of a new thread interleaving in each iteration of a test.
It follows in the footsteps of related tools such as CHESS \cite{chess},
but also provides several important features for use in an educational setting,
such as heuristics to handle unusual student solutions, kernel-level testing, and user-friendly progress reports and HTML debugging traces.

@@

{\bf Example.}
Figure \ref{fig:paradise}(a) shows an example concurrency bug. %which we will use as a running example.
The code implements a producer/consumer program, in which one thread sends work to the other through a shared list,
synchronized with a mutex and condition variable.
However, because the consumer code only checks once whether work exists before blocking,
%multiple
simultaneous consumers suffer
%as shown in Figure \ref{fig:paradise}(b),
a segfault when one receives a null \x{work} pointer. %and tries to dereference it

\definecolor{thread1}{RGB}{87,172,255}
\definecolor{thread2}{RGB}{255,201,102}
\definecolor{thread3}{RGB}{255,128,160}

\begin{figure}[t]
	\begin{tabular}{l|l}
	Consumer thread & Producer thread \\
	\hline
\begin{lstlisting}
mutex_lock(mx);
if (!work_exists())
  cond_wait(cvar,mx);
work = dequeue();
mutex_unlock(mx);
access(work->data);
\end{lstlisting}
	&
\begin{lstlisting}
mutex_lock(mx);

enqueue(work);
signal(cvar);

mutex_unlock(mx);
\end{lstlisting}
	\end{tabular}
	{\bf (a}) A buggy producer/consumer program, in which the consumer thread only checks \x{work_exists()} once.
	\vspace{1em}
	\\
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%	\begin{tabular}{l|l|l}
%	\cellcolor{thread1} {\bf Thread 1} & \cellcolor{thread2} {\bf Thread 2} & \cellcolor{thread3} {\bf Thread 3} \\
%	\hline
%\begin{lstlisting}
%lock();
%if (!work())
%  wait();
%\end{lstlisting}
%& & \\
%&
%\begin{lstlisting}
%lock();
%enqueue();
%signal();
%unlock();
%\end{lstlisting}
%& \\
%& &
%\begin{lstlisting}
%lock();
%dequeue();
%unlock();
%\end{lstlisting}
%\\
%\begin{lstlisting}
%dequeue();
%unlock();
%// SIGSEGV
%\end{lstlisting}
%& &
%	\begin{center}
%	{\bf (b)} Interleaving with two consumer threads. The second bypasses \x{cond_wait()}, causing the first to receive \x{NULL}. %in which one receives \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\includegraphics[width=0.48\textwidth]{execution-tree.pdf}
	{\bf (b)} The state space of all interleavings forms an execution tree. %The interleaving in (b) is the rightmost branch.
	The bug is observed when the second consumer thread bypasses \x{cond_wait()}, causing the first to receive \x{NULL}.
	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
	\caption{Example concurrency bug.}
	\label{fig:paradise}
\end{figure}

@@

The branching possibilities arising from each preemption point produce an execution tree, or state space,
as shown in Figure \ref{fig:paradise}(b).
%As these state spaces are exponentially-sized,
Landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
to mitigate the exponential explosion of these state spaces.
% TODO make sure this gets mentioned later, if not here
%and state space estimation \cite{estimation} to predict which
% TODO CAMREADY: cite oopsla
% In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.
This reduction requires only tracing the memory accesses of each thread,
so we are able to perform it behind-the-scenes, without consuming any student attention.

@@

@@

\newcommand\hilight[2]{\color{#1}#2\color{black}\xspace}
\definecolor{orange}{RGB}{192,96,0}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}
\definecolor{commentblue}{RGB}{0,0,192}

\begin{figure*}[t]
	\hilight{brickred}{{\bf TID 4 was killed by a page fault! (Faulting eip: 0x1000294; addr: 0x00000000)}} (53 interleavings tested)
	\\
	\begin{tabular}{|l|l|l|}
		\hline
		\multicolumn{1}{|c|}{\bf TID 4} &
		\multicolumn{1}{c|}{\bf TID 5} &
		\multicolumn{1}{c|}{\bf TID 6} \\
		\hline
		\hline
		0x010003d5 \hilight{olivegreen}{\x{deschedule}} (syscalls.c:7) & & \\
		0x010011a4 \hilight{olivegreen}{\x{cond_wait}} (cond.c:21) & & \\
		0x010001e0 \hilight{olivegreen}{\x{consumer}} (main.c:30) & & \\
		0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) & & \\
		\hline
		& 0x010034f5 \hilight{olivegreen}{\x{mutex_unlock}} (mutex.c:19) & \\
		& 0x01000037 \hilight{olivegreen}{\x{producer}} (main.c:74) & \\
		& 0x01000100 \hilight{olivegreen}{\x{main}} (main.c:88) & \\
		\hline
		& & 0x010034f5 \hilight{olivegreen}{\x{mutex_unlock}} (mutex.c:19) \\
		& & 0x0100020a \hilight{olivegreen}{\x{consumer}} (main.c:33) \\
		& & 0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) \\
		\hline
		0x01000294 \hilight{olivegreen}{\x{access}} (main.c:17) & & \\
		0x01000216 \hilight{olivegreen}{\x{consumer}} (main.c:34) & & \\
		0x01000259 \hilight{olivegreen}{\x{child_thread}} (main.c:42) & & \\
		\hline
	\end{tabular}
	\caption{Example preemption trace provided by Landslide as debugging output for the bug in Figure~\ref{fig:paradise}.}
	\label{fig:trace}
\end{figure*}

\subsection{Instrumentation}

For the thread library project, Landslide instruments student implementations automatically.
We provide a custom reference kernel binary equipped with the kernel instrumentations (described below)
so Landslide can track threads' lifecycles. %the lifecycle of threads.
The userspace \x{malloc} and \x{assert} are also provided in the basecode, and tracked similarly.
Landslide's data-race analysis must understand the locking discipline,
so it searches the student's code for functions named \x{mutex_lock} and \x{mutex_unlock} (as mandated by the assignment handout),
and assumes they at least provide mutual exclusion\footnote{
	To relax this assumption, we provide a special test case for mutexes,
	in which Landslide may preempt and search for data races in the mutex implementation itself.
}.
All other thread library components are treated as unknown, and subject to preemption anywhere.

To test \pebbles kernels directly, they must be annotated to inform Landslide of certain important concurrency events.
Because the kernel project leaves internal code structure as an open-ended design decision,
Landslide cannot search for specific function names. % like it does for P2s.
Instead, we provide a set of functions named with the prefix \x{tell_landslide}
to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
Users must also list all core scheduler, context switch, and timer handling routines, so Landslide can ignore their memory accesses for the sake of DPOR
(otherwise, threads would always conflict with each other and DPOR would achieve no reduction).
